---
title: "dplyr functions"
author: "Lab ___"
date: "Directions: Answer the bolded questions."
output: 
  slidy_presentation: 
    css: ../../IDSLabCSSRev.css
    mathjax: ../../extras/mathjax-local/MathJax.js
  word_document:
  ioslides_presentation: default
---

```{r, echo = FALSE, message = FALSE}
library(dplyr)
library(mobilizr)
data(cdc)
```

## Overview

<ul>

<li>`dplyr` is an R package for data manipulation.</li>

<li>**Data manipulation** refers to editing the structure of a data frame, such as adding or deleting variables, adding or deleting observations, etc.</li>

<li>The advantage of `dplyr` is that it consists of five main functions with intuitive names: `mutate`, `select`, `filter`, `summarize`, and `arrange`. These five verbs, combined with the **pipe** (`%>%`) and `group_by` make for data manipulation code that is easy to interpret.</li>

<li>For additional resources, run the following code:</li>

```{r, eval = FALSE}
vignette("dplyr")
```

</ul>

## The pipe

<ul>

<li>Often in `dplyr`, functions need to be **nested**.</li>

<li>**Nesting** is when a function takes another function as an argument.</li>

<li>One of the most important elements of `dplyr` is the **pipe** (`%>%`). The pipe makes long sequences of nested functions much more readable.</li> 

<li>Let's see how the pipe works on the next slide.</li>

</ul>

## How does the pipe work?

<ul>

<li>Suppose that for some reason, we are interested in finding the square root of the absolute value of the cosine of the number 2.</li>

<li>One way to compute this value is the following line of code:</li>

```{r}
sqrt(abs(cos(2)))
```

<li>To see the order in which functions are performed on the number 2, you have to read from inside to outside.</li>

<li>Using the pipe, we could instead write</li>

```{r}
2 %>% 
  cos() %>%
  abs() %>%
  sqrt()
```

<li>The pipe takes the object on the left side and makes it the first argument of the function on the right side. Chaining pipes together allows you to read the order in which functions are performed from left to right.</li>

<li>Notice how there is a new line after every pipe. This is not strictly necessary, but it is common practice and makes the code more readable.</li>

<li>**Which snippet is more readable?**</li>

</ul>

## select

<ul>

<li>`select` is used to subset columns (head is used to print only the first 6 rows).</li>

```{r}
cdc %>%
  select(height, weight) %>%
  head()
```

<li>To select a range of consecutive variables, use a colon (:).</li>

```{r}
cdc %>%
  select(race:seat_belt) %>%
  head()
```

<li>To take the complement of a set of variables, use an exclamation mark (!).</li>

```{r}
# The outermost function called is names, to get the variable names of the new data frame
# Notice that height is not present in the list of names
cdc %>%
  select(!height) %>%
  names()
```

<li>We can also use `where` to select those columns that satisfy a logical condition.</li>

```{r}
# height and weight are the only numeric columns
cdc %>%
  select(where(is.numeric)) %>%
  head()
```

</ul>

## filter

<ul>

<li>`filter` is used to subset rows.</li>

```{r}
cdc %>%
  filter(height > 1.8, weight > 90) %>%
  select(height, weight) %>%
  head()
```

</ul>

## mutate

<ul>

<li>`mutate` is used to add or edit the columns of a dataframe. You can add new variables that are functions of existing variables:</li>

```{r}
cdc %>%
  mutate(bmi = weight / height^2) %>%
  select(height, weight, bmi) %>%
  head()
```

<li>You can also edit existing variables:</li>

```{r}
cdc %>%
  mutate(height = height * 39.3701) %>%
  select(height) %>%
  head()
```

</ul>

## summarize

<ul>

<li>`summarize` reduces multiple values down to a single summary:</li>

```{r, warning = FALSE}
cdc %>%
  summarize(mean_height = mean(height), mean_weight = mean(weight))
```

</ul>

## group_by

<ul>

<li>Most data operations are done on groups defined by variables. `group_by` takes a dataframe and converts it into a grouped dataframe where operations are performed "by group". `group_by` will almost always come directly before a call to `summarize` or `mutate`.</li>

```{r, warning = FALSE}
cdc %>%
  group_by(gender) %>%
  summarize(mean_height = mean(height))
```

```{r, warning = FALSE}
cdc %>%
  group_by(gender) %>%
  mutate(mean_height = mean(height)) %>%
  select(gender, height, mean_height)
```

<li>You can group by multiple columns:</li>

```{r, warning = FALSE}
cdc %>%
  group_by(gender, depressed) %>%
  summarize(mean_height = mean(height))
```

</ul>

## arrange

<ul>

<li>`arrange` orders the rows of a data frame by the values of selected columns.</li>

```{r, warning = FALSE}
cdc %>%
  arrange(height) %>%
  select(height, weight) %>%
  head()
```

```{r, warning = FALSE}
cdc %>%
  arrange(height, weight) %>%
  select(height, weight) %>%
  head()
```

```{r, warning = FALSE}
cdc %>%
  arrange(weight, height) %>%
  select(height, weight) %>%
  head()
```

</ul>